---
title: "IDS investigation worksheet"
author: "the RegRession Rebels: Kenny, Jake, Matthew, Sarah & Alex"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(readxl)
library(tidymodels)

```


```{r load-data}
# Cleaning the data sets

diagnosis <- read_csv("data/mental-health-inpatient-activity-diagnosis-trend.csv", show_col_types = FALSE)

deprivation <- read_csv("data/mental-health-inpatient-activity-deprivation.csv", show_col_types = FALSE)

age_and_sex <- read_csv("data/mental-health-inpatient-activity-age-sex.csv", show_col_types = FALSE)

icd_10_codes <- read_xlsx("data/icd10-lookup.xlsx")

diagnosis_BUT_READABLE <- diagnosis %>% mutate(icd_categories = case_when(
ICD10Codes == "F00 - F09 or Chapter G and second position F00" ~ "Organic, including symptomatic, mental disorders", 
ICD10Codes == "F10 - F19" ~ "Mental & behavioural disorders due to psychoactive substance use", 
ICD10Codes == "F20 - F29" ~ "Schizophrenia, schizotypal, & delusional disorders", 
ICD10Codes == "F30 - F39" ~ "Mood (affective) disorders", 
ICD10Codes == "F40 - F49" ~ "Neurotic, stress-related, & somatoform disorders", 
ICD10Codes == "F60 - F69" ~ "Disorders of adult personality & behaviour", 
ICD10Codes == "F50 - F59, F70 - F99" ~ "Other selected diagnoses principally affecting children & young people", 
ICD10Codes == "Non-F codes" ~ "Unrecognized diagnosis", 
.default = ICD10Codes
)) %>% select(!ICD10Codes)

diagnosis_BUT_READABLE <- diagnosis_BUT_READABLE %>% mutate(smr_translated = case_when(
Dataset == "SMR00" ~ "Outpatients",
Dataset == "SMR01" ~ "General/Acute Inpatients/Day Cases (including Geriatric Long Stay Inpatients)",
Dataset == "SMR02" ~ "Maternity Inpatients/Day Cases",
Dataset == "SMR04" ~ "Mental Health Inpatients/Day Cases",
.default = Dataset
)) %>% select(!Dataset)

diagnosis_BUT_READABLE <- diagnosis_BUT_READABLE %>% 
  select(
    FinancialYear, HBT, DischargeCount, DischargeRates, PatientsCount,
    PatientsRates,HospitalResidentsCount, HospitalResidentsRates,
    icd_categories, smr_translated)

diagnosis_BUT_READABLE <- diagnosis_BUT_READABLE %>% mutate(hbt_translated = case_when(
HBT == "S92000003" ~ "Scotland",
HBT == "S08000015" ~ "Ayrshire and Arran",
HBT == "S08000016" ~ "Borders",
HBT == "S08000017" ~ "Dumfries and Galloway",
HBT == "S08000018" ~ "Fife",
HBT == "S08000019" ~ "Forth Valley",
HBT == "S08000020" ~ "Grampian",
HBT == "S08000021" ~ "Greater Glasgow and Clyde",
HBT == "S08000022" ~ "Highland",
HBT == "S08000023" ~ "Lanarkshire",
HBT == "S08000024" ~ "Lothian",
HBT == "S08000025" ~ "Orkney",
HBT == "S08000026" ~ "Shetland",
HBT == "S08000027" ~ "Tayside",
HBT == "S08000028" ~ "Western Isles",
HBT == "S08000029" ~ "Fife",
HBT == "S08000030" ~ "Tayside",
HBT == "S08000031" ~ "Greater Glasgow and Clyde",
HBT == "S08000032" ~ "Lanarkshire",
.default = HBT
)) %>% select(!HBT)

deprivation_BUT_READABLE <- deprivation %>% mutate(smr_translated = case_when(
Dataset == "SMR00" ~ "Outpatients",
Dataset == "SMR01" ~ "General/Acute Inpatients/Day Cases (including Geriatric Long Stay Inpatients)",
Dataset == "SMR02" ~ "Maternity Inpatients/Day Cases",
Dataset == "SMR03" ~ "Scottish Index of Deprivation",
Dataset == "SMR04" ~ "Mental Health Inpatients/Day Cases",
.default = Dataset
)) %>% select(!Dataset)

deprivation_BUT_READABLE <- deprivation_BUT_READABLE %>% mutate(hbr_translated = case_when(
HBR == "S92000003" ~ "Scotland",
HBR == "S08000015" ~ "Ayrshire and Arran",
HBR == "S08000016" ~ "Borders",
HBR == "S08000017" ~ "Dumfries and Galloway",
HBR == "S08000018" ~ "Fife",
HBR == "S08000019" ~ "Forth Valley",
HBR == "S08000020" ~ "Grampian",
HBR == "S08000021" ~ "Greater Glasgow and Clyde",
HBR == "S08000022" ~ "Highland",
HBR == "S08000023" ~ "Lanarkshire",
HBR == "S08000024" ~ "Lothian",
HBR == "S08000025" ~ "Orkney",
HBR == "S08000026" ~ "Shetland",
HBR == "S08000027" ~ "Tayside",
HBR == "S08000028" ~ "Western Isles",
HBR == "S08000029" ~ "Fife",
HBR == "S08000030" ~ "Tayside",
HBR == "S08000031" ~ "Greater Glasgow and Clyde",
HBR == "S08000032" ~ "Lanarkshire",
.default = HBR
)) %>% select(!HBR)

deprivation_BUT_READABLE <- deprivation_BUT_READABLE %>% 
  select(
    FinancialYear, hbr_translated, smr_translated, DischargeCount, DischargeRates, PatientsCount,
    PatientsRates,HospitalResidentsCount, HospitalResidentsRates, SIMD)

age_and_sex_BUT_READABLE <- age_and_sex %>% mutate(hbr_translated = case_when(
HBR == "S92000003" ~ "Scotland",
HBR == "S08000015" ~ "Ayrshire and Arran",
HBR == "S08000016" ~ "Borders",
HBR == "S08000017" ~ "Dumfries and Galloway",
HBR == "S08000018" ~ "Fife",
HBR == "S08000019" ~ "Forth Valley",
HBR == "S08000020" ~ "Grampian",
HBR == "S08000021" ~ "Greater Glasgow and Clyde",
HBR == "S08000022" ~ "Highland",
HBR == "S08000023" ~ "Lanarkshire",
HBR == "S08000024" ~ "Lothian",
HBR == "S08000025" ~ "Orkney",
HBR == "S08000026" ~ "Shetland",
HBR == "S08000027" ~ "Tayside",
HBR == "S08000028" ~ "Western Isles",
HBR == "S08000029" ~ "Fife",
HBR == "S08000030" ~ "Tayside",
HBR == "S08000031" ~ "Greater Glasgow and Clyde",
HBR == "S08000032" ~ "Lanarkshire",
.default = HBR
)) %>% select(!HBR)

age_and_sex_BUT_READABLE <- age_and_sex_BUT_READABLE %>% 
  select(
    FinancialYear, AgeGroup, Sex, hbr_translated, DischargeCount, DischargeRates, PatientsCount,
    PatientsRates)
```

```{r visualisations 1}
# Initial Visualizations
# age group
age_and_sex_BUT_READABLE %>%
  group_by(FinancialYear, AgeGroup) %>%
  summarise(Sum_DischargeCount = sum(DischargeCount, na.rm = TRUE), .groups = "keep") %>%
  
  ggplot(mapping = aes(x = AgeGroup, y = Sum_DischargeCount, fill = FinancialYear)) +
  geom_col() +
  labs(title = "Histogram of Discharge Count by Age Group and Financial Year",
       x = "AgeGroup",
       y = "Sum of DischargeCount") +
  theme_minimal()


age_and_sex_BUT_READABLE %>%
  group_by(FinancialYear, AgeGroup) %>%
  summarise(Sum_DischargeCount = sum(DischargeCount, na.rm = TRUE), .groups = "keep") %>%
  
  ggplot(mapping = aes(x = AgeGroup, y = Sum_DischargeCount, color = FinancialYear)) +
  geom_jitter(size = 2) +
  labs(title = "Scatter Plot of Discharge Count by Age Group and Financial Year",
       x = "AgeGroup",
       y = "Sum of DischargeCount",
       color = "FinancialYear") +
  theme_minimal()

```


```{r visualisations 2}
#sex
age_and_sex_BUT_READABLE %>%
  group_by(FinancialYear, Sex) %>%
  summarise(Sum_PatientsCount = sum(PatientsCount, na.rm = TRUE), .groups = "keep") %>%
  
  ggplot(mapping = aes(x = Sex, y = Sum_PatientsCount, fill = FinancialYear)) +
  geom_col() +
  labs(title = "Histogram of Patients Count by Age Group and Financial Year",
       x = "Sex",
       y = "Sum of Patients Count") +
  theme_minimal()


  age_and_sex_BUT_READABLE %>%
  group_by(FinancialYear, Sex) %>%
  summarise(Sum_DischargeCount = sum(DischargeCount, na.rm = TRUE), .groups = "keep") %>%
  
  ggplot(mapping = aes(x = Sex, y = Sum_DischargeCount, fill = FinancialYear)) +
  geom_col() +
  labs(title = "Histogram of Discharge Count by Age Group and Financial Year",
       x = "Sex",
       y = "Sum of Discharge Count") +
  theme_minimal()
```

```{r visualisations 3}
#icd_categories
diagnosis_BUT_READABLE %>%
  group_by( FinancialYear) %>%
  summarise(Sum_PatientsCount = sum(PatientsCount, na.rm = TRUE), .groups = "keep") %>%
  
  ggplot(mapping = aes(x = FinancialYear, y = Sum_PatientsCount)) +
  geom_point() +
  labs(title = "Histogram of Patients Count by Category",
       x = "Financial Year",
       y = "Sum of Patients Count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 6))


diagnosis_BUT_READABLE %>%
  group_by( FinancialYear) %>%
  summarise(Sum_DischargeCount = sum(DischargeCount, na.rm = TRUE), .groups = "keep") %>%
  
  ggplot(mapping = aes(x = FinancialYear, y = Sum_DischargeCount)) +
  geom_point() +
  labs(title = "Histogram of Discharge Count by Category",
       x = "Financial Year",
       y = "Sum of Discharge numbers") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 6))
```

```{r visualisations 4}
#icd_categories
diagnosis_BUT_READABLE %>%
  group_by( icd_categories) %>%
  summarise(Sum_PatientsCount = sum(PatientsCount, na.rm = TRUE), .groups = "keep") %>%
  
  ggplot(mapping = aes(x = icd_categories, y = Sum_PatientsCount)) +
  geom_col() +
  labs(title = "Histogram of Patients Count by Category",
       x = "icd_categories",
       y = "Sum of Patients Count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 6))


diagnosis_BUT_READABLE %>%
  group_by( icd_categories) %>%
  summarise(Sum_DischargeCount = sum(DischargeCount, na.rm = TRUE), .groups = "keep") %>%
  
  ggplot(mapping = aes(x = icd_categories, y = Sum_DischargeCount)) +
  geom_col() +
  labs(title = "Histogram of Discharge Count by Category",
       x = "icd categories",
       y = "Sum of Discharge Count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 6))
```

```{r visualisations 5}
# hbt_translated
diagnosis_BUT_READABLE %>%
  group_by( hbt_translated) %>%
  summarise(Sum_PatientsCount = sum(PatientsCount, na.rm = TRUE), .groups = "keep") %>%
  
  ggplot(mapping = aes(x = hbt_translated, y = Sum_PatientsCount)) +
  geom_col() +
  labs(title = "Histogram of Patients Count by Region",
       x = "Region",
       y = "Sum of Patients Count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 6))


diagnosis_BUT_READABLE %>%
  group_by(hbt_translated) %>%
  summarise(Sum_DischargeCount = sum(DischargeCount, na.rm = TRUE), .groups = "keep") %>%
  
  ggplot(mapping = aes(x = hbt_translated, y = Sum_DischargeCount)) +
  geom_col() +
  labs(title = "Histogram of Discharge Count by Region",
       x = "Region",
       y = "Sum of Discharge Count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 6))
```


```{r visualisations 6}
#deprivation
ggplot(deprivation_BUT_READABLE, aes(x = as.factor(SIMD), y = PatientsCount)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Patients Count vs SIMD",
       x = "SIMD",
       y = "Patients Count") +
  theme_minimal() 

ggplot(deprivation_BUT_READABLE, aes(x = SIMD, y = PatientsRates)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  labs(title = "Scatterplot with Linear Regression Line: SIMD vs Patients Rates",
       x = "SIMD",
       y = "Patients Rate") +
  theme_minimal()





```
