---
title: "IDS investigation worksheet"
author: "the RegRession Rebels: Kenny, Jake, Matthew, Sarah & Alex"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(readxl)


```


```{r load-data}
diagnosis <- read_csv("data/mental-health-inpatient-activity-diagnosis-trend.csv", show_col_types = FALSE)

deprivation <- read_csv("data/mental-health-inpatient-activity-deprivation.csv", show_col_types = FALSE)

icd_10_codes <- read_xlsx("data/icd10-lookup.xlsx")

diagnosis_BUT_READABLE <- diagnosis %>% mutate(icd_categories = case_when(
ICD10Codes == "F00 - F09 or Chapter G and second position F00" ~ "Organic, including symptomatic, mental disorders", 
ICD10Codes == "F10 - F19" ~ "Mental & behavioural disorders due to psychoactive substance use", 
ICD10Codes == "F20 - F29" ~ "Schizophrenia, schizotypal, & delusional disorders", 
ICD10Codes == "F30 - F39" ~ "Mood (affective) disorders", 
ICD10Codes == "F40 - F49" ~ "Neurotic, stress-related, & somatoform disorders", 
ICD10Codes == "F60 - F69" ~ "Disorders of adult personality & behaviour", 
ICD10Codes == "F50 - F59, F70 - F99" ~ "Other selected diagnoses principally affecting children & young people", 
ICD10Codes == "Non-F codes" ~ "Unrecognized diagnosis", 
.default = ICD10Codes
)) %>% select(!ICD10Codes)

diagnosis_BUT_READABLE <- diagnosis_BUT_READABLE %>% mutate(smr_translated = case_when(
Dataset == "SMR00" ~ "Outpatients",
Dataset == "SMR01" ~ "General/Acute Inpatients/Day Cases (including Geriatric Long Stay Inpatients)",
Dataset == "SMR02" ~ "Maternity Inpatients/Day Cases",
Dataset == "SMR04" ~ "Mental Health Inpatients/Day Cases",
.default = Dataset
)) %>% select(!Dataset)

diagnosis_BUT_READABLE <- diagnosis_BUT_READABLE %>% 
  select(
    FinancialYear, HBT, DischargeCount, DischargeRates, PatientsCount,
    PatientsRates,HospitalResidentsCount, HospitalResidentsRates,
    icd_categories, smr_translated)
```



