---
title: "Change in Patient Count Over Time with Respect to Different Regions and SIMD"
author: "the RegRession Rebels <br> Kenny Chen, Matthew Brown, Jake Leung, Aleksandra Iwaszkiewicz, Sarah De Giuli"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
# Add any other libraries here



```


```{r load-data, include=FALSE}
# load your data here and any pre-processing/cleaning 
#  that you need for your report.




```
How do discharge numbers in Scotland change over time?
  - try fitting a suitable model to quantify the change in the numbers and predict how they might          behave in the future

3 additional means of analysis:
- how they compare to patient numbers over time
- How they are affected by the deprivation index


## Research Question

Describe your research question here. Include why this question might be of interest. 

The research question aims to identify whether there exists a correlation between how numbers of patients and numbers of patients discharged have changed over time. Additionally, the investigation analyzes the SIMD score, the index of deprivation, to determine the existence of a correlation and potential impact of the assigned deprivation index of an area on the mental health of its inhabitants, reflected and discerned on the basis of the discharge numbers and patient counts. 

*insert specific methods of accomplishing the aim, explain the means undertaken to answer the rq* 

This topic of investigation is worth exploring due to its potential to understanding how the approach to treating mental health has developed over the years, from 1997 to modern day. This could function as a basis for making predictions about the future numbers of patients in mental health institutions, and therefore, form conclusions about the efficiency and success of treating mental health in Scotland. While this may be of interest to the investigation, it is important to remember that a number of assumptions are being made, which will be stated and explained appropriately throughout the process that is being documented. As insightful as the data used for the investigation is, this is a topic of great importance and extent, so it is worth bearing in mind that this exploration should not be treated as a whole and complete analysis, more so a beginning to the conversation that offers initial conclusions prompting further questions and allowing room to delve into the specifics and intricacies of mental health related data analysis. 

## Data

Briefly describe your dataset here - include a description of the key variables you will be investigating. 
Include a citation for your data. 
See http://libraryguides.vu.edu.au/c.php?g=386501&p=4347840 for guidance on proper citation for data sets. 
If you got your data off the web, make sure to state the retrieval date.

Our data set is about mental heath data in Scotland. It covers the period from 1977 to 2022. It provides helpful data for our study such as SIMD(Scottish Index of Multiple Deprivation), financial year, patient count and discharge count. 

NHS Scotland 
2018
Mental Health Inpatient Activity
Electronic Dataset
https://www.opendata.nhs.scot/dataset/mental-health-inpatient-activity accessed on Friday 20th October 2023

## Findings

The write-up of your project and findings go here. 

Think of this as the text of your presentation with some extra detail to cover what there was not time to discuss in the presentation. This might include any assumptions you made when doing your analysis, any limitations of the work you have done or any ideas you have for future work. Feel free to split this section into subsections to make it easier to read. 

The length should be roughly 1,500 words. 
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio. Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good! 
The addin will ignore code chunks and only count the words in prose. If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.





## References

List any references here. You should, at a minimum, list your data source.

https://www.opendata.nhs.scot/dataset/mental-health-inpatient-activity accessed on Friday 20th October 2023
