---
title: "Project title"
author: "the RegRession Rebels <br> Kenny Chen, Matthew Brown, Jake Leung, Aleksandra Iwaszkiewicz, Sarah De Giuli"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
# Add any other libraries here



```


```{r load-data, include=FALSE}
# load your data here and any pre-processing/cleaning 
#  that you need for your report.




```
How do discharge numbers in Scotland change over time?
  - try fitting a suitable model to quantify the change in the numbers and predict how they might          behave in the future

3 additional means of analysis:
- how they compare to patient numbers over time
- how they are distributed by region
- How they are affected by the deprivation index


## Research Question

Describe your research question here. Include why this question might be of interest. 

To see if there is a correlation between how patient/discharge count has changed over time and your SIMD score

This is of interest as it allows us to understand how mental health has changed over time, and potentially make predictions about mental health in the future. We also have the oppertunity to analyse the impact of deprivation on mental heath and show the effects deprivation has on people in Scotland.

## Data

Briefly describe your dataset here - include a description of the key variables you will be investigating. 
Include a citation for your data. 
See http://libraryguides.vu.edu.au/c.php?g=386501&p=4347840 for guidance on proper citation for datasets. 
If you got your data off the web, make sure to state the retrieval date.

Our data set is about mental heath data in Scotland. It covers the period from 1977 to 2022. It provides helpful data for our study such as SIMD(Scottish Index of Multiple Deprivation), financial year, patient count and discharge count. 

NHS Scotland 
2018
Mental Health Inpatient Activity
Electronic Dataset
https://www.opendata.nhs.scot/dataset/mental-health-inpatient-activity accessed on Friday 20th October 2023

## Findings

The write-up of your project and findings go here. 

Think of this as the text of your presentation with some extra detail to cover what there was not time to discuss in the presentation. This might include any assumptions you made when doing your analysis, any limitations of the work you have done or any ideas you have for future work. Feel free to split this section into subsections to make it easier to read. 

The length should be roughly 1,500 words. 
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio. Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good! 
The addin will ignore code chunks and only count the words in prose. If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.





## References

List any references here. You should, at a minimum, list your data source.

https://www.opendata.nhs.scot/dataset/mental-health-inpatient-activity accessed on Friday 20th October 2023
